Density-dependent selection and the limits of relative fitness

### 2018-03-20 manuscript.tex
l

The preceding arguments are well-known, and appear in various forms in standard textbooks (e.g. \citealt{crow_1970,begon_1990,nagylaki_1992,otto_2011,ewens_2004}). However, the constant relative fitness idealization of population genetics need not \textit{only} apply when selection is weak (or when selection in genuinely density-independent). Since this idealization is so powerful, it is important to understand the specifics of how and when it fails when selection is not weak, or $N$ is not stable. For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}. Are we compelled to switch to a more ecologically-detailed model of selection based on Malthusian parameters or birth/death rates in this important model system? And if we make this switch, how much ecological detail do we need?

The models of density-regulated growth mentioned so far are all missing an aspect of selection with major ramifications for the constant-$s$ description of selection: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles. Although similar ideas featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}), few attempts have been made to explicitly model density-dependent selection in the presence of reproductive excesses. \cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a discussion of Haldane's ``cost of selection'' \citep{haldane_1957}, but used an unusual model of competition which was defined for at most two different types. In more complex models with detailed age structure (as opposed to juvenile vs adult), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding, but otherwise the view of density dependence is surprisingly similar to that found in the above models without age structure \citep[Chap. 4]{charlesworth_1994}. For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp. 148]{charlesworth_1994}. As a result, the role of reproductive excesses in justifying the use of constant relative finesses is still largely verbal.

Here we investigate the population ecology of relative fitness using a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess. We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery already incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). However, the classic lottery also assumes constant $N$. Our first task is to relax the constant-$N$ assumption of the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').

>>>>

However, the preceding arguments do not imply that the constant relative fitness idealization of population genetics \textit{only} applies when selection is weak and $N$ is stable (or when selection is actually density-independent). This idealization is powerful, and so it is important to understand the specifics of how it succeeds or fails when selection is not weak, or $N$ is not stable. For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}. Are we compelled to switch to a more ecologically-detailed model of selection based on Malthusian parameters or birth/death rates in this important model system? And if we make this switch, how much ecological detail do we need?

Here we argue that the simplified models of density-regulated growth mentioned above give a narrow view of density-dependent selection that is misleading in some important ways. These ultimately derive from their failure to account for the presence of ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles. Although similar ideas featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}), few attempts have been made to explicitly model density-dependent selection in the presence of reproductive excesses. \cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a discussion of Haldane's ``cost of selection'' \citep{haldane_1957}, but used an unusual model of competition which was defined for at most two different types. In more complex models with detailed age structure (as opposed to juvenile vs adult), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding, but otherwise the view of density dependence is surprisingly similar to that found in the models mentioned above, which have no age structure \citep[Chap. 4]{charlesworth_1994}. For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp. 148]{charlesworth_1994}. As a result, the extent to which reproductive excesses justify the use of constant relative finesses is unclear beyond heuristic verbal arguments.

We investigate the population ecology of relative fitness using a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess. We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery already incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). However, the classic lottery also assumes constant $N$. Our first task is to relax the constant-$N$ assumption of the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
